---
title: "ReproducableReserarch Assignment1"
author: "Michael Shanley"
date: "March 15, 2015"
output: html_document
---

##Introduction

##It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

##This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.</p>

##<p><b>Lets Get 'R ready!</b></p>

##Setting echo to True so that everyone can see the code

##<p>This is an R HTML document. When you click the <b>Knit HTML</b> button a web page will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:</p>

```{r}
library(knitr)
opts_chunk$set(echo = TRUE, results = 'hold')
```

##<p>Load required libraries</p>
##<p>Lets use green plots since it is March.</p>

```{r}
library(ggplot2)
```

##<p>Load the data.</p>

##Note: It is assumed that the file activity.csv is in the current working directory. File can be downloaded from here:

```{r}
setwd("~/Desktop/ReproduceableResearch")
rdata <- read.csv('activity.csv', header = TRUE, sep = ",", colClasses=c("numeric", "character", "numeric"))
```

##Lets format some stuff!

##Dates should be dates, intervals go to factor.

```{r}
rdata$date <- as.Date(rdata$date, format = "%Y-%m-%d")
rdata$interval <- as.factor(rdata$interval)
```


##Check it out:

```{r}
str(rdata)
```
## 'data.frame':    17568 obs. of  3 variables:
##  $ steps   : num  NA NA NA NA NA NA NA NA NA NA ...
##  $ date    : Date, format: "2012-10-01" "2012-10-01" ...
##  $ interval: Factor w/ 288 levels "0","5","10","15",..: 1 2 3 4 5 6 7 8 9 10 ...

What is mean total number of steps taken per day?

##<p>Lets get rid of missing values.  Whats with the data collection anywho?</P.

##<p>Total steps per day.</p>

```{r}
steps_per_day <- aggregate(steps ~ date, rdata, sum)
colnames(steps_per_day) <- c("date","steps")
head(steps_per_day)
```
##         date steps
## 1 2012-10-02   126
## 2 2012-10-03 11352
## 3 2012-10-04 12116
## 4 2012-10-05 13294
## 5 2012-10-06 15420
## 6 2012-10-07 11015
Now we make a histogram of the total number of steps taken per day, plotted with appropriate bin interval.
```{r}
ggplot(steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Steps Taken per Day", 
             x = "Steps per Day", y = "Number of times a day(Count)") + theme_bw() 
```

Calculate the mean and median of the number of steps per day.
```{r}
steps_mean   <- mean(steps_per_day$steps, na.rm=TRUE)
steps_median <- median(steps_per_day$steps, na.rm=TRUE)
```

The mean is 10766.189 and median is 10765.

##<p>What is the average daily activity pattern?</p>

##We calculate the aggregation of steps by intervals of 5-minutes and convert the intervals as integers and save them in a data frame called steps_per_interval.

```{r}
steps_per_interval <- aggregate(rdata$steps, 
                                by = list(interval = rdata$interval),
                                FUN=mean, na.rm=TRUE)
#convert to integers to plot

steps_per_interval$interval <- 
        as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```
Plot the time series of average number of steps taken versus the 5-minute intervals:
```{r} 
ggplot(steps_per_interval, aes(x=interval, y=steps)) +   
        geom_line(color="green", size=1) +  
        labs(title="Average Daily Activity Pattern", x="Interval", y="Number of steps") +  
        theme_bw()
```
##plot of chunk plot_time_series

Find the 5-minute interval with the maximum number of steps:
```{r} 
max_interval <- steps_per_interval[which.max(  
        steps_per_interval$steps),]
```
The 835th interval has 206 steps.

Imputing missing values:

1. Total number of missing values:

The total number of missing values in steps can be calculated using is.na() method to check whether the value is mising or not and then summing the logical vector.

```{r}
missing_vals <- sum(is.na(rdata$steps))
```
The total number of missing values are 2304.

2. Strategy for filling in all of the missing values in the dataset

To populate missing values, we choose to replace them with the mean value at the same interval across days. In most of the cases the median is a better centrality measure than mean, but in our case the total median is not much far away from total mean, and probably we can make the mean and median meets.

##We create a function na_fill(data, pervalue) which the data arguement is the rdata data frame and pervalue arguement is the steps_per_interval data frame.
```{r}
na_fill <- function(data, pervalue) {
        na_index <- which(is.na(data$steps))
        na_replace <- unlist(lapply(na_index, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        fill_steps <- data$steps
        fill_steps[na_index] <- na_replace
        fill_steps
}

rdata_fill <- data.frame(  
        steps = na_fill(rdata, steps_per_interval),  
        date = rdata$date,  
        interval = rdata$interval)
str(rdata_fill)
```
## 'data.frame':    17568 obs. of  3 variables:
##  $ steps   : num  1.717 0.3396 0.1321 0.1509 0.0755 ...
##  $ date    : Date, format: "2012-10-01" "2012-10-01" ...
##  $ interval: Factor w/ 288 levels "0","5","10","15",..: 1 2 3 4 5 6 7 8 9 10 ...
We check that are there any missing values remaining or not

```{r}
sum(is.na(rdata_fill$steps))
```
## [1] 0
Zero output shows that there are no missing values exist.

3. A histogram of the total number of steps taken each day

Now let us plot a histogram of the daily total number of steps taken, plotted with a bin interval of 1000 steps, after filling missing values.

```{r}
fill_steps_per_day <- aggregate(steps ~ date, rdata_fill, sum)
colnames(fill_steps_per_day) <- c("date","steps")
```
##plot
```{r}
ggplot(fill_steps_per_day, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1000) + 
        labs(title="Steps Taken per Day", 
             x = "Steps per Day", y = "Number of times in a day(Count)") + theme_bw() 
```

Calculate and report the mean and median total number of steps taken per day.

```{r}
steps_mean_fill   <- mean(fill_steps_per_day$steps, na.rm=TRUE)
steps_median_fill <- median(fill_steps_per_day$steps, na.rm=TRUE)
```
The mean is 10766.189 and median is 10766.189.

Do these values differ from the estimates from the first part of the assignment?

A wee bit.

Before:

Mean : 10766.189
Median: 10765

AfterL
Mean : 10766.189
Median: 10766.189

After correcting for missing data the Median and the Mean match.

What is the impact of imputing missing data on the estimates of the total daily number of steps?

It brings the mean and the median into alignment.

Predictions remain the same with or without the missing values, however the high end goes up.

Are there differences in activity patterns between weekdays and weekends?

We do this comparison with the table with filled-in missing values.
1. Augment the table with a column that indicates the day of the week
2. Subset the table into two parts - weekends (Saturday and Sunday) and weekdays (Monday through Friday).
3. Tabulate the average steps per interval for each data set.
4. Plot the two data sets side by side for comparison.

```{r}
weekdays_steps <- function(data) {
    weekdays_steps <- aggregate(data$steps, by=list(interval = data$interval),
                          FUN=mean, na.rm=T)
    # convert to integers for plotting
    weekdays_steps$interval <- 
            as.integer(levels(weekdays_steps$interval)[weekdays_steps$interval])
    colnames(weekdays_steps) <- c("interval", "steps")
    weekdays_steps
}

data_by_weekdays <- function(data) {
    data$weekday <- 
            as.factor(weekdays(data$date)) # weekdays
    weekend_data <- subset(data, weekday %in% c("Saturday","Sunday"))
    weekday_data <- subset(data, !weekday %in% c("Saturday","Sunday"))

    weekend_steps <- weekdays_steps(weekend_data)
    weekday_steps <- weekdays_steps(weekday_data)

    weekend_steps$dayofweek <- rep("weekend", nrow(weekend_steps))
    weekday_steps$dayofweek <- rep("weekday", nrow(weekday_steps))

    data_by_weekdays <- rbind(weekend_steps, weekday_steps)
    data_by_weekdays$dayofweek <- as.factor(data_by_weekdays$dayofweek)
    data_by_weekdays
}

data_weekdays <- data_by_weekdays(rdata_fill)
```

Next, plot the average number of steps taken per 5-minute interval across weekdays and weekends:

```{r}
ggplot(data_weekdays, aes(x=interval, y=steps)) + 
        geom_line(color="violet") + 
        facet_wrap(~ dayofweek, nrow=2, ncol=1) +
        labs(x="Interval", y="Number of steps") +
        theme_bw()
plot of chunk plot_weekdays
```

Surprisingly, the weekday had the highest number of steps taken.  Hoever, overall activity was greater on the weekend.  The workday seems to be more uniform, for the most part, while the weekend is more active.



